Positivity-Preserving Adaptive Runge--Kutta Methods

### 2020-04-29 nusslein_vsrp/latex/overview.tex
move revised results for stratospheric reaction system into the main file

\section{Conclusion} \label{sec:conclusion}
It is possible to adapt the weights to enforce positivity for RKMs that are not positivity preserving. 
One main limitation is that the resulting order has to be lower than the number of stages.
An error approximation for this method was given.  
The region of absolute stability is altered by changing the weights. This effect can be predicted or controlled. 
Used with explicit methods the positivity for some test problems could be recovered.  
Because the time step size is limited by the stability it is only useful for a small interval of time steps. 
The adaptive method is mainly interesting for diagonally implicit methods.
The times step size is not limited by stability. 
Also, the cost of solving the LP is not a crucial factor.  
If the negative values occurring are not too large, which can be expected for most computations, adapting the weights is a potential way to ensure positivity.

>>>>

\subsection{Stiff problem with adaptive step size II}

Here, we consider the stratospheric reaction problem of
\cite{sandu2001positive}, which models the reactions of the substances
in the concentration vector $u = [O^{1D}, O, O_3, O_2, NO, NO_2]$.
This ODE has two linear invariants
\begin{align}
  m_O^T u &= \text{const}, & m_O &= [1,1,3,2,1,2]^T, \\
  m_N^T u &= \text{const}, & m_N &= [0,0,0,0,1,1]^T,
\end{align}
which describe the conservation of the total mass of oxygen and nitrogen,
respectively.

The reaction system
\begin{equation}
\label{eq:stratospheric}
\begin{aligned}
  \frac{\mathrm d}{\mathrm d t} O^{1D} &= r_5 - r_6 - r_7, \\
  \frac{\mathrm d}{\mathrm d t} O      &= 2 r_1 - r_2 + r_3 - r_4 + r_6 - r_9 + r_{10} - r_{11}, \\
  \frac{\mathrm d}{\mathrm d t} O_3    &= r_2 - r_3 - r_4 - r_5 - r_7 - r_8, \\
  \frac{\mathrm d}{\mathrm d t} O_2    &= -r_1 - r_2 + r_3 + 2 r_4 + r_5 + 2r_7 + r_8 + r_9, \\
  \frac{\mathrm d}{\mathrm d t} NO     &= -r_8 + r_9 + r_{10} - r_{11}, \\
  \frac{\mathrm d}{\mathrm d t} NO_2   &= r_8 - r_9 - r_{10} + r_{11},
\end{aligned}
\end{equation}
with time $t$ in seconds is given by the reaction rates
\begin{equation}
\begin{aligned}
  r_{1}  &= k_1\, O_2,          &  k_{1}  &= \num{2.643e-10} \sigma^3, &
  r_{2}  &= k_2\, O\, O_2,      &  k_{2}  &= \num{8.018e-17}, \\
  r_{3}  &= k_3\, O_3,          &  k_{3}  &= \num{6.120e-4} \sigma, &
  r_{4}  &= k_4\, O_3\, O,      &  k_{4}  &= \num{1.567e-15}, \\
  r_{5}  &= k_5\, O_3,          &  k_{5}  &= \num{1.070e-3} \sigma^2, &
  r_{6}  &= k_6\, M\, O^{1D},   &  k_{6}  &= \num{7.110e-11}, \\
  r_{7}  &= k_7\, O^{1D}\, O_3, &  k_{7}  &= \num{1.200e-10}, &
  r_{8}  &= k_8\, O_3\, NO,     &  k_{8}  &= \num{6.062e-15}, \\
  r_{9}  &= k_9\, NO_2\, O,     &  k_{9}  &= \num{1.069e-11}, &
  r_{10} &= k_{10}\, NO_2,      &  k_{10} &= \num{1.289e-2} \sigma, \\
% \stepcounter{equation}\tag{\theequation}
  r_{11} &= k_{11}\, NO\, O,    &  k_{11} &= \num{1.0e-8},
\end{aligned}
\end{equation}
where $M  = \num{8.120e16}$ and
\begin{align}
    T &= (t/3600) \bmod 24,
    \quad T_r = 4.5,
    \quad T_s = 19.5 \\
    \sigma(T) &=
    \begin{cases}
    0.5+0.5 \cos\Bigl(\pi \Bigl|\frac{(2T-T_r-T_s)}{(T_s-T_r)}\Bigr| \frac{(2T-T_r-T_s)}{(T_s-T_r)} \Bigr)       & \quad \text{if } T_r \leq T \leq T_s,\\
    0  & \quad \text{otherwise}.
    \end{cases}
\end{align}

The initial conditions are
\begin{equation}
  u(t_0) = [
    \num{9.906e+1},
    \num{6.624e08},
    \num{5.326e11},
    \num{1.697e16},
    \num{4.000e6},
    \num{1.093e9}
  ]^T.
\end{equation}
The system was normalized internally such that $\forall n\colon u_n(t_0) = 1$
for the computation to achieve a suitable error estimation.
The system is solved in the time from $t_{0} = \SI{12}{h}$ to $t = \SI{84}{h}$
using the Im-Euler~3 extrapolation method with free adaptation and step size
control.

\begin{figure}
\centering
\includegraphics[width=0.9\textwidth]{plots/Stratospheric_time.pdf}
\caption{Numerical approximation of stratospheric reaction system
         \eqref{eq:stratospheric}.}
\label{fig:Stratospheric_time}
\end{figure}

The results are shown in Figure~\ref{fig:Stratospheric_time}.
The adapted solution is close to the reference solution obtained
with the unadapted Im-Euler~3 extrapolation method and a higher accuracy.
For this solution, 249 steps were computed; two of these were rejected due to
a violation of the error bound.
More details are shown in Figure~\ref{fig:Stats_Strat}. The rejected steps are drawn with thick crosses.
The step size $\dt$ undergoes multiple sudden changes due to the explicit dependence
on time of the problem.
The minimum values before and after the adaptation are also shown for all steps where the initial values were negative.
This was only the case for some time intervals. 25 steps exhibited negative values.
Almost all of them were very close to zero and the adaptation did only show a small improvement.
The smallest value of the solution is $\min(u) = \num{-1.59e-11}$.
The used weights are also very close to the original weights, except of the steps that were rejected anyway due to a violation of the tolerance.
The change of the two linear invariants are shwon in the last two subplots. Both are preserved within roundoff error.
Due to the internal normalization the round off error is amplified to the order of $10$ for $m_O^Tu$.

\todo{TODO: Plot relative change $(m_O^T u - m_O^T u(t_0)) / m_O^T u(t_0)$
instead of the absolute change of the linear invariants?}

\begin{figure}
\centering
\includegraphics[width=0.85\textwidth]{plots/Stratospheric_stepsize,b.pdf}
\caption{Different statistics of computation of the stratospheric reaction system
         \eqref{eq:stratospheric}.}
\label{fig:Stats_Strat}
\end{figure}

\section{Conclusion} \label{sec:conclusion}

It is possible to adapt the weights to enforce positivity for RKMs that are not positivity preserving.
One main limitation is that the resulting order has to be lower than the number of stages.
An error approximation for this method was given.
The region of absolute stability is altered by changing the weights. This effect can be predicted or controlled.
Used with explicit methods the positivity for some test problems could be recovered.
Because the time step size is limited by the stability it is only useful for a small interval of time steps.
The adaptive method is mainly interesting for diagonally implicit methods.
The times step size is not limited by stability.
Also, the cost of solving the LP is not a crucial factor.
If the negative values occurring are not too large, which can be expected for most computations, adapting the weights is a potential way to ensure positivity.
